---
title: "Pole Vault"
author: "Lainey Rose"
date: "2023-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First we need to import the data from the website, 'alltimeathletics.com'

```{r}
library(tidyverse)
library(readr)
womensPoleVault <- read.table("womenspolevault.txt", sep = '%', quote = "\"")
mensPoleVault <- read.table("menspolevault.txt", sep = '%', quote = "\"")
```
Next, I will need to create a table for the data to be stored in and later visualized. However, I tried to do this simply using read.delim, and there were many errors. There were some issues with how the data was separated, as well as a lack of column labels. To fix this, I created a vector for the column labels and separated each value by at least two spaces, like so:

```{r}
newWomens <- tidyr::separate(womensPoleVault, V1, c('EMPTY', 'RANK', 'HEIGHT', 
  'NAME', 'COUNTRY', 'BIRTHDATE', 'NUM', 'LOCATION', 'DATE'), sep = "\\s{2,}", ) %>%
  select(-EMPTY)
head(newWomens)
View(newWomens)

newMens <- tidyr::separate(mensPoleVault, V1, c('EMPTY', 'RANK', 'HEIGHT', 
  'NAME', 'COUNTRY', 'BIRTHDATE', 'NUM', 'LOCATION', 'DATE'), sep = "\\s{2,}", ) %>%
  select(-EMPTY)
head(newMens)
View(newMens)
```

Now that I have loaded the data for both the men's and women's events, I will start by simplifying the data so that it only has what I will need for my analysis.

```{r}
womensSimplified <- newWomens %>% select(DATE, BIRTHDATE) %>% drop_na(DATE, BIRTHDATE)

mensSimplified <- newMens %>% select(DATE, BIRTHDATE) %>% drop_na(DATE, BIRTHDATE) 
```

I can now create a new table that has two additonal columns that only state the year, so that I can create an algorithm that strips the periods from both dates. 

```{r}
womenRevised <- womensSimplified %>% mutate(YEAR = substr(DATE, 7, 10)) %>% 
  mutate(BIRTHYEAR = substr(BIRTHDATE, 7, 10))
womenRevised$YEAR <- as.numeric(as.character(womenRevised$YEAR))
womenRevised$BIRTHYEAR <- as.numeric(as.character(womenRevised$BIRTHYEAR))
View(womenRevised)
sapply(womenRevised, class)

menRevised <- mensSimplified %>% mutate(YEAR = substr(DATE, 7, 10)) %>% 
  mutate(BIRTHYEAR = substr(BIRTHDATE, 7, 10))
menRevised$YEAR <- as.numeric(as.character(menRevised$YEAR))
womenRevised$BIRTHYEAR <- as.numeric(as.character(womenRevised$BIRTHYEAR))
sapply(menRevised, class)
View(menRevised)

```
I now have to find the age of each individual by subtracting the date the event occurred by their birthdate minus 1900.

```{r}
#womenRevised %>% mutate(AGE = )
```


Now time for visualizing the data, using a histogram.

```{r}
library(ggplot2)
ggplot(womensSimplified, aes(x = HEIGHT)) + geom_histogram()
       
ggplot(mensSimplified, aes(x = HEIGHT)) + geom_histogram()

```

