---
title: "Pole Vault"
author: "Lainey Rose"
date: "2023-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load data

```{r}
library(tidyverse)
library(readr)
library(ggplot2)
womensPoleVault <- read.table("womenspolevault.txt", sep = '%', quote = "\"")
newWomens <- tidyr::separate(womensPoleVault, V1, c('EMPTY', 'RANK', 'HEIGHT', 
  'NAME', 'COUNTRY', 'BIRTHDATE', 'NUM', 'LOCATION', 'DATE'), sep = "\\s{2,}", ) %>%
  select(-EMPTY)
head(newWomens)
mensPoleVault <- read.table("menspolevault.txt", sep = '%', quote = "\"")
newMens <- tidyr::separate(mensPoleVault, V1, c('EMPTY', 'RANK', 'HEIGHT', 
  'NAME', 'COUNTRY', 'BIRTHDATE', 'NUM', 'LOCATION', 'DATE'), sep = "\\s{2,}", ) %>%
  select(-EMPTY)
head(newMens)


```

Now that I have loaded the data for both the men's and women's events, I need to see where mistakes were made.It looks there were many parsing errors made that could affect my data analysis. After a quick look at it, there are a few errors where the names seem to mess up the data, so I need to find a way to fix this whether that be manually or by writing more code.


I can also simplify the data since I only need to use a few of the columns. This might mean that the columns that were affected won't even be used. Let's start by simplifying the data:

```{r}
womensSimplified <- newWomens %>% select('HEIGHT', 'DATE')
mensSimplified <- newMens %>% select('HEIGHT', 'DATE')
```


Now time for visualizing the data, using a histogram.

```{r}
ggplot(womensSimplified, aes(x = 'DATE', y = 'HEIGHT')) + geom_point()
       
ggplot(mensSimplified, aes(x = 'DATE', y = 'HEIGHT')) + geom_point()

```

